Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-01-06 supplement.tex
edits to fst discussion

\subsection{$F_{ST}$ between coexisting populations over time}

$F_{ST}$ can be interpreted as a measure of population structure, given by the
proportion of genetic variance in a multi-population sample that can be
attributed to drift between populations in that sample \cite{Bhatia2013-sq}.
$F_{ST}$ depends on population divergence times as well as subsequent gene flow
and population sizes: migration between populations reduces $F_{ST}$ while
small effective population sizes can lead to large values. For each of the four
inferred demographic models presented here, we computed predicted $F_{ST}$ for
pairs of contemporaneous populations over time. For sampled populations,
predicted $F_{ST}$ can be compared to data (Fig.~\ref{fig:supp-FST-panels}). We
computed $F_{ST}$ as a ratio of expected (or average) variances, using observed
pairwise diversity within and between populations as
\begin{equation}
    F_{ST}(i, j) = \frac{\E[2H_{i,j} - H_i - H_j]}{\E[2H_{i,j} + H_i + H_j]}.
    \label{eq:FST}
\end{equation}

In all inferred models, $F_{ST}$s between Neanderthals and the ancestors of
modern humans were much larger than between any pair of modern human
populations or their ancestors (stems). $F_{ST}$ between modern human
populations over the last 100 thousand years remained relatively low, at
$\sim0.1$ or less. In the continuous-migration model, the two stem populations
also were predicted to have low $F_{ST}$ within this range, as migration rates
were inferred to be high enough to maintain genetic similarity. In both the
merger-without-stem-migration and the merger-with-stem-migration, one of the
stem populations was inferred to have a short, severe bottleneck, which led to
increased $F_{ST}$ between stem populations during that time. However, this
period of relatively large $F_{ST}$ did not result in large $F_{ST}$ between
more recent pairs of populations due to their roughly similar levels of
contributions to the ancestors of modern humans. \aprcomment{We caution that
the severe bottleneck may be an artifact of the model parameterization?}

>>>>

\subsection{Divergence between coexisting populations over time}

We considered a few different measures of differentiation between populations  in four inferred models, starting with $F_{ST}.$ For each of the four
inferred demographic models presented here, we computed predicted $F_{ST}$ for
pairs of contemporaneous populations over time.

$F_{ST}$ represents the
proportion of genetic variance in a multi-population sample that can be
attributed to drift between populations in that sample (e.g., \cite{Bhatia2013-sq}).
$F_{ST}$ depends on population divergence times, gene flow,
and population sizes. For present-day sampled populations,
predicted $F_{ST}$ can be compared to data (Fig.~\ref{fig:supp-FST-panels}), while past estimates represent model predictions about the differentiation between past populations. We
computed $F_{ST}$ as a ratio of expected (or average) variances, using observed
pairwise diversity within and between populations as \sgcomment{WHich estimator is that? If this is the one recommended in Bhatia, cite?}
\begin{equation}
    F_{ST}(i, j) = \frac{\E[2H_{i,j} - H_i - H_j]}{\E[2H_{i,j} + H_i + H_j]}.
    \label{eq:FST}
\end{equation}
\sgcomment{where $H_{ij}$ is the average number of differences between one haploid copy of the genome from population $i$ and one from population $j$.}
In all inferred models, $F_{ST}$s between Neanderthals and the ancestors of
modern humans were much larger than between any pair of modern human
populations or their ancestors (stems). $F_{ST}$ between modern human
populations over the last $100ka$ remained relatively low, at
$\sim0.1$ or less. In the continuous-migration model, the two stem populations
also were predicted to have low $F_{ST}$ within this range, as migration rates
were inferred to be high enough to maintain genetic similarity. In both the
merger-without-stem-migration and the merger-with-stem-migration, one of the
stem populations was inferred to have a short, severe bottleneck, which led to
increased $F_{ST}$ between stem populations during that time. In the merger with stem migration, $F_{ST}$ reached a maximum value of $0.3$ between stems, intermediate between human-Neanderthal divergence ($0.54$ \sgcomment{check}) and the maximum $F_{ST}$ observed between contemporary human populations ($0.07$.)

Whereas the high $F_{ST}$ value between Neanderthal and humans reflects both elevated genetic differences between the genomes of modern humans and neanderthal and reduced diversity within the Neanderthal population (\sgcomment{Figure X}), the increased $F_{ST}$ between the stems is primarily due to an inferred bottleneck reducing stem diversity, while genetic differentiation between individuals from the two stems is within the range of genetic differentiation between (and even within) human populations. In other words, our best fit model describes two stems with one stems showing reduced diversity, but individuals across stems showing no more differentiation than two modern population.  \sgcomment{Previous paragraph could use rereading.}
